Cyclist video evidence of road traffic offences: Preliminary analysis of Operation Snap data and suggestions for a research agenda

### 2024-05-20 paper.qmd
Update paper.qmd
tweaked abstract

---
title: "Cyclist video evidence of road traffic offences: Preliminary analysis of Operation Snap data and suggestions for a research agenda"
bibliography: references.bib
editor:
  markdown: 
    wrap: sentence
author:
  - name: Graham Farrell
    corresponding: true
    affiliations:
      - name: University of Leeds
        department: School of Law
        address: The Liberty Building
        city: Leeds
        country: United Kingdom
        postal-code: LS2 9JT
    orcid: 0000-0002-3987-8457
  - name: Robin Lovelace
    affiliations:
      - name: University of Leeds
        department: Institute for Transport Studies
        address: University Road
        city: Leeds
        country: United Kingdom
        postal-code: LS2 9JT
      - name: Active Travel England
        # department: Data and Digital
        # address: West Offices Station Rise
        # city: York
        # country: United Kingdom
        # postal-code: YO1 6GA
    orcid: 0000-0001-5679-6536
  - name: Steve O'Hern
    affiliations:
      - name: University of Leeds
        department: Institute for Transport Studies
        address: University Road
        city: Leeds
        country: United Kingdom
        postal-code: LS2 9JT
      - name: Monash University Accident Research Centre
        address: Clayton, VIC 3800, Australia
    orcid: 0000-0002-7961-3875
keywords: 
  - Cycling
  - Hotspots 
  - Road safety
  - Video evidence
  - Dashcam
  - Near misses
date: last-modified
format:
  # pdf: default
  arxiv-pdf:
    keep-tex: true
linenumbers: true # Add (continuous) line numbers?
doublespacing: false # Double space the PDF output?
runninghead: "Preprint" # The text on the top of each page of the output
authorcols: true # Should authors be listed in a single column (default) or in multiple columns (`authorcols: true`)
---

@@

This study uses data from Operation Snap (OpSnap), the UK police’s national system to receive road users’ video evidence of road traffic offences.
Data from one police force area for three calendar years, from its launch in 2021 (N= 18,363 records), is analysed.
Of submitted cases, 34.4% were from cyclists (n=6,312) of which 69% resulted in an offence being registered, mostly for driving ‘without reasonable consideration to others’ or ‘without due care and attention’.
Three quarters (75.2%) of reported cases resulted in the recommended disposal of an educational course (including conditional offers), a quarter in no further action, and fewer than 1% for court appearance.
A research agenda using OpSnap data is outlined: comparative analysis of submission type, source, hotspot locations, and case disposal should be undertaken at national, regional, and local levels; research into disposal-related decision-making, video quality, and the role of supporting evidence should identify good practice; further study of video content, offence concentration (recidivism, repeat submitters of evidence, spatial hotspots) and case progression including court cases, holds significant potential to inform policy and practice locally, nationally and internationally.

@@

Dangerous and criminal driving are significant problems that take many forms [@simon1996; @corbett2003; @corbett2010].
Across Great Britain there were an estimated 1,711 fatalities resulting from a road traffic crash in 2022 and 135,480 causalities reported across all severities [@dft-xxxx].
Most injuries involve motor vehicle occupants, however when accounting for the distance travelled by road user groups, cyclists are over-represented in injury statistics (Department for Transport, 2023a).
Between 2004 and 2022, an average of 104 cyclists were killed and 4,212 were seriously injured each year in Britain according to police records (Department for Transport, 2023b).
Almost half of cyclist fatalities involved collision with a car, with 56% on rural roads (compared to 30% of traffic).
The 2023 report noted that “the most common contributory factor allocated to pedal cyclists in fatal or serious collisions (FSC) with another vehicle was ‘driver or rider failed to look properly’” (Department for Transport, 2023b).

```{=html}
<!-- The context for the study is that, between 2004 and 2022, an average of 104 cyclists were killed and 4,212 were seriously injured each year according to official records recorded by police forces across Great Britain [@departmentfortransport].
Almost half of cyclist fatalities involved collision with a car, with 56% on rural roads (compared to 30% of traffic).
The 2023 report noted that “the most common contributory factor allocated to pedal cyclists in fatal or serious collisions (FSC) with another vehicle was ‘driver or rider failed to look properly’” [@departmentfortransport].  -->
```

Cyclist crashes are underreported in police recorded datasets (Elvik & Mysen, 1999) and the extent of injuries sustained by cyclists may be higher when hospital-recorded cases are counted (Janstrup et al., 2016). Under-reporting in police datasets is likely to be greatest for minor injuries, while records are seldom kept when a collision or injury is avoided due to riders or drivers taking evasive actions (Ibrahim et al., 2021). These incidents are often referred to as “near-misses” (Ibrahim et al., 2021).

@@

Close passes are the most common type of near miss reported by cyclists, and are associated with collisions resulting in injury [@aldred2016]. A ‘close pass’ refers to when a vehicle passes too close to a cyclist, which is defined in the UK as less than 1.5 metres away at 30mph (50kph).  Close passes take different forms including the ‘punishment pass’ by angry drivers for a perceived slight such as causing the driver to slow down [@cubbin2024]. There is no specific law in the UK Road Traffic Act 1988 for driving too close to cyclist, but two are commonly applied for careless driving: RT88575- Drive without due care and attention; and RT99576 – Driver without reasonable consideration to others - and these are prominent in the analysis that follows.

This study highlights the potential of using open access data from Operation Snap, the UK police’s national system to receive road users’ video evidence of road traffic offences. Operation Snap, often referred to informally as ‘OpSnap’, was piloted by North Wales police in October 2016 and adopted by all Welsh forces by 20181. It is now in operation nationally across England and Wales, each police force offering its own submission portal for road users to submit video evidence. The nature of video submissions and the related expectations are summarised on the website of one Police and Crime Commissioner as follows:

- The secure form is for traffic offences, it is NOT for submitting footage of road traffic collisions, any other offences or for parking issues. 
- The car registration number of the offending vehicle must be clearly visible. 
- The public should be prepared to sign a witness statement and possibly give evidence in court. 
- Statements for OpSnap can only be accepted from persons aged 18 or over. If you are under 18 the incident should be reported by email to [email]2.

@@

Open access data from West Yorkshire Safer Roads – OpSnap was used for this study, the media submissions portal having opened in July 2020. The West Yorkshire Police (WYP) data used here spanned calendar years 2021, 2022 and 2023. For the year 2021 there was less than half the cases of either 2022 or 2023, which could reflect reduced road use during the COVID-19 pandemic and fewer people being aware of OpSnap when it commenced.

The dataset is a tabulation of cases submitted to the West Yorkshire Safer Roads OpSnap web portal. The terms ‘record’ and ‘case’ are used interchangeably here to refer to a record in the OpSnap database, each of which represents the separate submission of video evidence by a road user.

The portal allows members of the public to submit video footage of suspected traffic offences committed by motor vehicle drivers. Video footage is commonly recorded from on-board cameras.  For motor vehicles these cameras are typically mounted on or near the vehicle front dashboard, known as ‘dash cams’. One source examined GB Driving Licence data to find that by early 2024, close to a third of private and commercial vehicle in the UK had a dash cam (iCompario, 2024). For cyclists, footage is commonly recorded using helmet or handlebar-mounted cameras. The proportion of cyclists using these cameras is unknown but anecdotal evidence suggests increased usage, with many choosing to record rides in case an incident occurs.

Complainants upload footage and complete a short form that includes their personal details, the details of the vehicle involved including registration, make, model and colour, the location and time of the incident, and details of the camera used to record the footage. Only vehicle offences can be reported as the registration number of any offending vehicle is required and must be legible in footage. The open access data is a deidentified summary of submitted cases with informtion on: mode of transport of the person reporting, offender vehicle details (make, model, colour), offence code, recommended disposal, date of submission, district, and offence location. The offence location is typically a street name and town or city name, or an intersection, and examples included: ‘A58 Godley Road, Halifax’, ‘Keighley Road, Silsden’ , ‘Woodhouse Lane A660, Leeds’. For this study, approximate geolocations were obtained using Google API, restricting cases to within West Yorkshire. Further aspects of the data, their uses and limitations are discussed in what follows.

@@

Over 90 percent of offences reported by cyclists were for driving without consideration for other road users and without due care and attention. This is consistent with the phenomenon of ‘close passes’ in the cycling safety literature (Aldred, 2016; Cubbin et al., 2024). It suggests the OpSnap data holds the potential for further analysis to inform knowledge about the nature of close passes generally. For example, it should be possible to identify hotspot locations. As with much of the further research outlined here, this the potential to inform preventive policy and practice relating to police interventions, driver behaviour, the design of roads and roadside environments.

@@

Within the trend of increasing video submissions there were distinct seasonal patterns. Seasonal variation in cyclists’ submission reflect seasonal (weather and other) influences on cycling prevalence and duration. This suggests seasonal variation in preventive responses might be tailored to need. It is well established that crime is highly concentrated along whatever dimension is examined (Farrell & Pease, 2017). With respect to video evidence this will include recidivists (repeat offenders), repeat submitters of evidence (who may or may not be repeat victims/survivors), the concentration of incidents (close passes, near misses and crashes) at certain times and places, with different types of experiences concentrated among certain types of road users. For example, we hypothesise that close passes are more likely to be reported by cyclists and horse riders than vehicle drivers, both due to the nature of the road user interactions and the relative risk associated with the manoeuvre. As such, the types of incidents reported by vehicle drivers would be systematically different, indicating a need for different investigative and preventive approaches. Policing and preventions efforts focused on where problems are concentrated are more resource efficient – it is the foundation of problem-oriented policing (see e.g., (Laing, 2010; Scott & Maddox, 2009)- and research using video submissions holds the potential to analysis of problem concentrations.

Evidence submitted by cyclists was the focus here. However, Operation Snap data offer potential for analysis of data from other road user sources. Studies focused on specific types of road user will prove informative with respect to policy and practice. Dash cam submissions by motorists offers great potential to inform other aspects of road safety. Research into submissions by pedestrians, horse riders and motorcyclists should prove viable to improve the safety of these parties. There were over four hundred submissions by horse riders in West Yorkshire, which means there will be thousands nationwide. The British Horse Society^[
https://www.bhs.org.uk/
]
and others may be interested in this data being used to inform the safety of horses and riders. The British Motorcyclists Federation^[
https://www.britishmotorcyclists.co.uk/
]
may be interested in research promoting the safety of its members and in why motorcyclist submissions are relatively infrequent.

@@

The spatial analysis undertaken here was obliged to use data for ‘all’ video submissions. We were unable to separate the submissions by cyclists because, in the publicly-available data, locations could not be matched to individual cases: they were scrambled to ensure anonymity for GDPR purposes. An obvious next step, in the context of a secure research platform, would be spatial analysis for different types of road users, for different types of incidents, for incidents resulting in different disposals, and so on. Hotspots and spatial clustering are likely to vary by type of road user, type of incident, day or week, time of day, and so on.

>>>>

---
title: "Cyclist video evidence of road traffic offences: Preliminary analysis of Operation Snap data and suggestions for a research agenda"
bibliography: references.bib
editor:
  markdown: 
    wrap: sentence
author:
  - name: Graham Farrell
    corresponding: true
    affiliations:
      - name: University of Leeds
        department: School of Law
        address: The Liberty Building
        city: Leeds
        country: United Kingdom
        postal-code: LS2 9JT
    orcid: 0000-0002-3987-8457
  - name: Robin Lovelace
    affiliations:
      - name: University of Leeds
        department: Institute for Transport Studies
        address: University Road
        city: Leeds
        country: United Kingdom
        postal-code: LS2 9JT
      - name: Active Travel England
        # department: Data and Digital
        # address: West Offices Station Rise
        # city: York
        # country: United Kingdom
        # postal-code: YO1 6GA
    orcid: 0000-0001-5679-6536
  - name: Steve O'Hern
    affiliations:
      - name: University of Leeds
        department: Institute for Transport Studies
        address: University Road
        city: Leeds
        country: United Kingdom
        postal-code: LS2 9JT
      - name: Monash University Accident Research Centre
        address: Clayton, VIC 3800, Australia
    orcid: 0000-0002-7961-3875
keywords: 
  - Cycling
  - Hotspots 
  - Road safety
  - Video evidence
  - Dashcam
  - Near misses
date: last-modified
format:
  # pdf: default
  arxiv-pdf:
    keep-tex: true
linenumbers: true # Add (continuous) line numbers?
doublespacing: false # Double space the PDF output?
runninghead: "Preprint" # The text on the top of each page of the output
authorcols: true # Should authors be listed in a single column (default) or in multiple columns (`authorcols: true`)
editor_options: 
  markdown: 
    wrap: sentence
---

@@

This study uses data from Operation Snap (OpSnap), the UK police's national system to receive road users' video evidence of road traffic offences.
Data from one police force area for the three calendar years 2021 to 2023 (N= 18,363 records), is analysed.
Of submitted cases, 34.4% were from cyclists (n=6,312) of which 69% resulted in an offence being registered, mostly for driving 'without reasonable consideration to others' or 'without due care and attention'.
Three quarters (75.2%) of reported cases resulted in the recommended disposal of an educational course (including conditional offers), a quarter in no further action, and fewer than 1% for court appearance.
A research agenda using OpSnap data is outlined: comparative analysis of submission type, source, hotspot locations, and case disposal should be undertaken at national, regional, and local levels; research into disposal-related decision-making, video quality, and the role of supporting evidence should identify good practice; further study of video content, offence concentration (recidivism, repeat submitters of evidence, spatial hotspots) and case progression including court cases, holds significant potential to inform policy and practice locally, nationally and internationally.

@@

Dangerous and criminal driving are significant problems that take many forms [@simon1996; @corbett2003; @corbett2010].
Across Great Britain there were an estimated 1,711 fatalities resulting from a road traffic crash in 2022 and 135,480 causalities reported across all severities [@dft-xxxx].
Most injuries involve motor vehicle occupants, however when accounting for the distance travelled by road user groups, cyclists are over-represented in injury statistics (Department for Transport, 2023a).
Between 2004 and 2022, an average of 104 cyclists were killed and 4,212 were seriously injured each year in Britain according to police records (Department for Transport, 2023b).
Almost half of cyclist fatalities involved collision with a car, with 56% on rural roads (compared to 30% of traffic).
The 2023 report noted that "the most common contributory factor allocated to pedal cyclists in fatal or serious collisions (FSC) with another vehicle was 'driver or rider failed to look properly'" (Department for Transport, 2023b).

```{=html}
<!-- The context for the study is that, between 2004 and 2022, an average of 104 cyclists were killed and 4,212 were seriously injured each year according to official records recorded by police forces across Great Britain [@departmentfortransport].
Almost half of cyclist fatalities involved collision with a car, with 56% on rural roads (compared to 30% of traffic).
The 2023 report noted that “the most common contributory factor allocated to pedal cyclists in fatal or serious collisions (FSC) with another vehicle was ‘driver or rider failed to look properly’” [@departmentfortransport].  -->
```
Cyclist crashes are underreported in police recorded datasets (Elvik & Mysen, 1999) and the extent of injuries sustained by cyclists may be higher when hospital-recorded cases are counted (Janstrup et al., 2016).
Under-reporting in police datasets is likely to be greatest for minor injuries, while records are seldom kept when a collision or injury is avoided due to riders or drivers taking evasive actions (Ibrahim et al., 2021).
These incidents are often referred to as "near-misses" (Ibrahim et al., 2021).

@@

Close passes are the most common type of near miss reported by cyclists, and are associated with collisions resulting in injury [@aldred2016].
A 'close pass' refers to when a vehicle passes too close to a cyclist, which is defined in the UK as less than 1.5 metres away at 30mph (50kph).
Close passes take different forms including the 'punishment pass' by angry drivers for a perceived slight such as causing the driver to slow down [@cubbin2024].
There is no specific law in the UK Road Traffic Act 1988 for driving too close to cyclist, but two are commonly applied for careless driving: RT88575- Drive without due care and attention; and RT99576 -- Driver without reasonable consideration to others - and these are prominent in the analysis that follows.

This study highlights the potential of using open access data from Operation Snap, the UK police's national system to receive road users' video evidence of road traffic offences.
Operation Snap, often referred to informally as 'OpSnap', was piloted by North Wales police in October 2016 and adopted by all Welsh forces by 20181.
It is now in operation nationally across England and Wales, each police force offering its own submission portal for road users to submit video evidence.
The nature of video submissions and the related expectations are summarised on the website of one Police and Crime Commissioner as follows:

-   The secure form is for traffic offences, it is NOT for submitting footage of road traffic collisions, any other offences or for parking issues. 
-   The car registration number of the offending vehicle must be clearly visible. 
-   The public should be prepared to sign a witness statement and possibly give evidence in court. 
-   Statements for OpSnap can only be accepted from persons aged 18 or over. If you are under 18 the incident should be reported by email to \[email\]2.

@@

Open access data from West Yorkshire Safer Roads -- OpSnap was used for this study, the media submissions portal having opened in July 2020.
The West Yorkshire Police (WYP) data used here spanned calendar years 2021, 2022 and 2023.
For the year 2021 there was less than half the cases of either 2022 or 2023, which could reflect reduced road use during the COVID-19 pandemic and fewer people being aware of OpSnap when it commenced.

The dataset is a tabulation of cases submitted to the West Yorkshire Safer Roads OpSnap web portal.
The terms 'record' and 'case' are used interchangeably here to refer to a record in the OpSnap database, each of which represents the separate submission of video evidence by a road user.

The portal allows members of the public to submit video footage of suspected traffic offences committed by motor vehicle drivers.
Video footage is commonly recorded from on-board cameras.
For motor vehicles these cameras are typically mounted on or near the vehicle front dashboard, known as 'dash cams'.
One source examined GB Driving Licence data to find that by early 2024, close to a third of private and commercial vehicle in the UK had a dash cam (iCompario, 2024).
For cyclists, footage is commonly recorded using helmet or handlebar-mounted cameras.
The proportion of cyclists using these cameras is unknown but anecdotal evidence suggests increased usage, with many choosing to record rides in case an incident occurs.

Complainants upload footage and complete a short form that includes their personal details, the details of the vehicle involved including registration, make, model and colour, the location and time of the incident, and details of the camera used to record the footage.
Only vehicle offences can be reported as the registration number of any offending vehicle is required and must be legible in footage.
The open access data is a deidentified summary of submitted cases with informtion on: mode of transport of the person reporting, offender vehicle details (make, model, colour), offence code, recommended disposal, date of submission, district, and offence location.
The offence location is typically a street name and town or city name, or an intersection, and examples included: 'A58 Godley Road, Halifax', 'Keighley Road, Silsden' , 'Woodhouse Lane A660, Leeds'.
For this study, approximate geolocations were obtained using Google API, restricting cases to within West Yorkshire.
Further aspects of the data, their uses and limitations are discussed in what follows.

@@

Over 90 percent of offences reported by cyclists were for driving without consideration for other road users and without due care and attention.
This is consistent with the phenomenon of 'close passes' in the cycling safety literature (Aldred, 2016; Cubbin et al., 2024).
It suggests the OpSnap data holds the potential for further analysis to inform knowledge about the nature of close passes generally.
For example, it should be possible to identify hotspot locations.
As with much of the further research outlined here, this the potential to inform preventive policy and practice relating to police interventions, driver behaviour, the design of roads and roadside environments.

@@

Within the trend of increasing video submissions there were distinct seasonal patterns.
Seasonal variation in cyclists' submission reflect seasonal (weather and other) influences on cycling prevalence and duration.
This suggests seasonal variation in preventive responses might be tailored to need.
It is well established that crime is highly concentrated along whatever dimension is examined (Farrell & Pease, 2017).
With respect to video evidence this will include recidivists (repeat offenders), repeat submitters of evidence (who may or may not be repeat victims/survivors), the concentration of incidents (close passes, near misses and crashes) at certain times and places, with different types of experiences concentrated among certain types of road users.
For example, we hypothesise that close passes are more likely to be reported by cyclists and horse riders than vehicle drivers, both due to the nature of the road user interactions and the relative risk associated with the manoeuvre.
As such, the types of incidents reported by vehicle drivers would be systematically different, indicating a need for different investigative and preventive approaches.
Policing and preventions efforts focused on where problems are concentrated are more resource efficient -- it is the foundation of problem-oriented policing (see e.g., (Laing, 2010; Scott & Maddox, 2009)- and research using video submissions holds the potential to analysis of problem concentrations.

Evidence submitted by cyclists was the focus here.
However, Operation Snap data offer potential for analysis of data from other road user sources.
Studies focused on specific types of road user will prove informative with respect to policy and practice.
Dash cam submissions by motorists offers great potential to inform other aspects of road safety.
Research into submissions by pedestrians, horse riders and motorcyclists should prove viable to improve the safety of these parties.
There were over four hundred submissions by horse riders in West Yorkshire, which means there will be thousands nationwide.
The British Horse Society[^1] and others may be interested in this data being used to inform the safety of horses and riders.
The British Motorcyclists Federation[^2] may be interested in research promoting the safety of its members and in why motorcyclist submissions are relatively infrequent
.

[^1]:  https://www.bhs.org.uk/

[^2]:  https://www.britishmotorcyclists.co.uk/

@@

The spatial analysis undertaken here was obliged to use data for 'all' video submissions.
We were unable to separate the submissions by cyclists because, in the publicly-available data, locations could not be matched to individual cases: they were scrambled to ensure anonymity for GDPR purposes.
An obvious next step, in the context of a secure research platform, would be spatial analysis for different types of road users, for different types of incidents, for incidents resulting in different disposals, and so on.
Hotspots and spatial clustering are likely to vary by type of road user, type of incident, day or week, time of day, and so on.
